---
title: "Classification Trees and Ensemble Methods"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = T, message = FALSE, results='hide', warning=FALSE}
library(mlbench)
library(caret)
library(rpart)
library(rpart.plot)
library(party)
library(partykit)
library(randomForest)
library(ranger)
library(gbm)
library(plotmo)
library(pdp)
library(pROC)
library(lime)
```

We use the Pima Indians Diabetes Database (used in L6.Rmd) for illustration. The data contain 768 observations and 9 variables. The outcome is a binary variable `diabetes`. 

```{r}
data(PimaIndiansDiabetes)
dat <- PimaIndiansDiabetes
dat$diabetes <- factor(dat$diabetes, c("pos", "neg"))

set.seed(1)
rowTrain <- createDataPartition(y = dat$diabetes,
                                p = 0.75,
                                list = FALSE)
```

# Classification trees

## Using `rpart`

```{r}
set.seed(1)
tree1 <- rpart(formula = diabetes~., data = dat,
               subset = rowTrain, 
               control = rpart.control(cp = 0))

cpTable <- printcp(tree1)
plotcp(tree1)
minErr <- which.min(cpTable[,4])

# minimum cross-validation error
tree2 <- prune(tree1, cp = cpTable[minErr,1])
rpart.plot(tree2)
```

## Using `caret`

### CART

```{r, fig.height=2.8, fig.width=3.5, fig.show='hold'}
ctrl <- trainControl(method = "repeatedcv",
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE)

set.seed(1)
rpart.fit <- train(diabetes~., dat, 
                   subset = rowTrain,
                   method = "rpart",
                   tuneGrid = data.frame(cp = exp(seq(-6,-3, len = 20))),
                   trControl = ctrl,
                   metric = "ROC")
ggplot(rpart.fit, highlight = TRUE)
rpart.plot(rpart.fit$finalModel)
```

### CIT

```{r}
set.seed(1)
ctree.fit <- train(diabetes~., dat, 
                   subset = rowTrain,
                   method = "ctree",
                   tuneGrid = data.frame(mincriterion = 1-exp(seq(-2, -1, length = 20))),
                   metric = "ROC",
                   trControl = ctrl)
ggplot(ctree.fit, highlight = TRUE)
```

```{r, fig.width=15, fig.height=6}
plot(ctree.fit$finalModel)
```

```{r}
rpart.pred <- predict(tree2, newdata = dat[-rowTrain,])[,1]

rpartc.pred <- predict(rpart.fit, newdata = dat[-rowTrain,],
                       type = "prob")[,1]
ctree.pred <- predict(ctree.fit, newdata = dat[-rowTrain,],
                       type = "prob")[,1]
```

# Random forests and boosting

```{r}
set.seed(1)
bagging <- randomForest(diabetes~., dat[rowTrain,],
                   mtry = 8)

set.seed(1)
rf <- randomForest(diabetes~., dat[rowTrain,],
                   mtry = 3)

# fast implementation
set.seed(1)
rf2 <- ranger(diabetes~., dat[rowTrain,],
              mtry = 3, probability = TRUE) 


rf.pred <- predict(rf, newdata = dat[-rowTrain,], type = "prob")[,1]
rf2.pred <- predict(rf2, data = dat[-rowTrain,], type = "response")$predictions[,1]
```


```{r}
dat2 <- dat
dat2$diabetes <- as.numeric(dat$diabetes == "pos")

set.seed(1)
bst <- gbm(diabetes~., dat2[rowTrain,],
           distribution = "adaboost",
           n.trees = 1000, 
           interaction.depth = 2,
           shrinkage = 0.005,
           cv.folds = 10)

nt <- gbm.perf(bst, method = "cv")
nt
```

## Grid search using `caret`

### Random forests

```{r}
# Try more if possible
rf.grid <- expand.grid(mtry = 1:6,
                       splitrule = "gini",
                       min.node.size = 1:6)
set.seed(1)
rf.fit <- train(diabetes~., dat, 
                subset = rowTrain,
                method = "ranger",
                tuneGrid = rf.grid,
                metric = "ROC",
                trControl = ctrl)

ggplot(rf.fit, highlight = TRUE)

rf.pred <- predict(rf.fit, newdata = dat[-rowTrain,], type = "prob")[,1]
```

### Boosting

#### Binomial loss

```{r}
gbmB.grid <- expand.grid(n.trees = c(2000,3000,4000),
                        interaction.depth = 1:6,
                        shrinkage = c(0.001,0.003,0.005),
                        n.minobsinnode = 1)
set.seed(1)
# Binomial loss function
gbmB.fit <- train(diabetes~., dat, 
                 subset = rowTrain, 
                 tuneGrid = gbmB.grid,
                 trControl = ctrl,
                 method = "gbm",
                 distribution = "bernoulli",
                 metric = "ROC",
                 verbose = FALSE)

ggplot(gbmB.fit, highlight = TRUE)

gbmB.pred <- predict(gbmB.fit, newdata = dat[-rowTrain,], type = "prob")[,1]
```

#### AdaBoost

```{r}
gbmA.grid <- expand.grid(n.trees = c(2000,3000,4000),
                        interaction.depth = 1:6,
                        shrinkage = c(0.001,0.003,0.005),
                        n.minobsinnode = 1)
set.seed(1)
# Adaboost loss function
gbmA.fit <- train(diabetes~., dat, 
                 subset = rowTrain, 
                 tuneGrid = gbmA.grid,
                 trControl = ctrl,
                 method = "gbm",
                 distribution = "adaboost",
                 metric = "ROC",
                 verbose = FALSE)

ggplot(gbmA.fit, highlight = TRUE)

gbmA.pred <- predict(gbmA.fit, newdata = dat[-rowTrain,], type = "prob")[,1]
```


```{r}
resamp <- resamples(list(rf = rf.fit, 
                         gbmA = gbmA.fit,
                         gbmB = gbmB.fit,
                         rpart = rpart.fit,
                         ctree = ctree.fit))
summary(resamp)
```

## Understanding your models
### Variable importance

```{r}
set.seed(1)
rf2.final.per <- ranger(diabetes~., dat[rowTrain,], 
                        mtry = 3, 
                        min.node.size = 5,
                        splitrule = "gini",
                        importance = "permutation",
                        scale.permutation.importance = TRUE) 

barplot(sort(ranger::importance(rf2.final.per), decreasing = FALSE), 
        las = 2, horiz = TRUE, cex.names = 0.7,
        col = colorRampPalette(colors = c("cyan","blue"))(8))

set.seed(1)
rf2.final.imp <- ranger(diabetes~., dat[rowTrain,], 
                        mtry = 3, splitrule = "gini",
                        min.node.size = 5,
                        importance = "impurity") 

barplot(sort(ranger::importance(rf2.final.imp), decreasing = FALSE), 
        las = 2, horiz = TRUE, cex.names = 0.7,
        col = colorRampPalette(colors = c("cyan","blue"))(8))
```



```{r}
summary(gbmA.fit$finalModel, las = 2, cBars = 19, cex.names = 0.6)
```

### PDP 

```{r}
pdp.rf <- rf.fit %>% 
  partial(pred.var = "glucose", 
          grid.resolution = 100,
          prob = TRUE) %>%
  autoplot(rug = TRUE, train = dat[rowTrain,]) +
  ggtitle("Random forest") 

pdp.gbm <- gbmA.fit %>% 
  partial(pred.var = "glucose", 
          grid.resolution = 100,
          prob = TRUE) %>%
  autoplot(rug = TRUE, train = dat[rowTrain,]) +
  ggtitle("Boosting") 

grid.arrange(pdp.rf, pdp.gbm, nrow = 1)
```

### ICE

```{r}
ice1.rf <- rf.fit %>% 
  partial(pred.var = "glucose", 
          grid.resolution = 100,
          ice = TRUE,
          prob = TRUE) %>%
  autoplot(train = dat[rowTrain,], alpha = .1) +
  ggtitle("Random forest, non-centered") 

ice2.rf <- rf.fit %>% 
  partial(pred.var = "glucose", 
          grid.resolution = 100,
          ice = TRUE,
          prob = TRUE) %>%
  autoplot(train = dat[rowTrain,], alpha = .1, 
           center = TRUE) +
  ggtitle("Random forest, centered") 

ice1.gbm <- gbmA.fit %>% 
  partial(pred.var = "glucose", 
          grid.resolution = 100,
          ice = TRUE,
          prob = TRUE) %>%
  autoplot(train = dat[rowTrain,], alpha = .1) +
  ggtitle("Boosting, non-centered") 

ice2.gbm <- gbmA.fit %>% 
  partial(pred.var = "glucose", 
          grid.resolution = 100,
          ice = TRUE,
          prob = TRUE) %>%
  autoplot(train = dat[rowTrain,], alpha = .1, 
           center = TRUE) +
  ggtitle("Boosting, centered") 

grid.arrange(ice1.rf, ice2.rf, ice1.gbm, ice2.gbm,
             nrow = 2, ncol = 2)
```

### Explain your prediction

```{r, warning=FALSE}
new_obs <- dat[-rowTrain,-9][1:2,]
explainer.gbm <- lime(dat[rowTrain,-9], gbmA.fit)
explanation.gbm <- explain(new_obs, explainer.gbm, n_features = 8,
                           labels = "pos")
plot_features(explanation.gbm)
```

```{r, warning=FALSE}
explainer.rf <- lime(dat[rowTrain,-9], rf.fit)
explanation.rf <- explain(new_obs, explainer.rf, n_features = 8,
                          labels = "pos")
plot_features(explanation.rf)
```


# Test data performance

```{r}
roc.rpart <- roc(dat$diabetes[-rowTrain], rpart.pred)
roc.rpartc <- roc(dat$diabetes[-rowTrain], rpartc.pred)
roc.ctree <- roc(dat$diabetes[-rowTrain], ctree.pred)
roc.rf <- roc(dat$diabetes[-rowTrain], rf.pred)
roc.gbmA <- roc(dat$diabetes[-rowTrain], gbmA.pred)
roc.gbmB <- roc(dat$diabetes[-rowTrain], gbmB.pred)


plot(roc.rpart)
plot(roc.rpartc, add = TRUE, col = 2)
plot(roc.ctree, add = TRUE, col = 3)
plot(roc.rf, add = TRUE, col = 4)
plot(roc.gbmA, add = TRUE, col = 5)
plot(roc.gbmB, add = TRUE, col = 6)


auc <- c(roc.rpart$auc[1], roc.rpartc$auc[1], roc.ctree$auc[1],
         roc.rf$auc[1], roc.gbmA$auc[1], roc.gbmB$auc[1])

modelNames <- c("rpart","rpart_caret","ctree","rf","gbmA","gbmB")
legend("bottomright", legend = paste0(modelNames, ": ", round(auc,3)),
       col = 1:6, lwd = 2)
```
